---
jupyter:
  jupytext:
    formats: ipynb,Rmd
    text_representation:
      extension: .Rmd
      format_name: rmarkdown
      format_version: '1.2'
      jupytext_version: 1.11.0
  kernelspec:
    display_name: R
    language: R
    name: ir
---

## **Mediation Analysis**
---


## Input


### Libraries

```{r}
# Library
spsm <- suppressPackageStartupMessages 
spsm(library(dplyr))
spsm(library(tidyverse))
spsm(library(FactoMineR))
spsm(library(ggplot2))
spsm(library(factoextra))
spsm(library(corrplot))
spsm(library(mediation))
spsm(library(mosaic))
spsm(library(kableExtra))
spsm(library(IRdisplay))
spsm(library(vioplot))
```

### Ext. Functions

```{r}
# Function
source("../../utils/getData.R")
source("../../utils/getPCs.R")
source("../../utils/getPermut.R")
source("../../utils/getMediationTable.R")
source("../../utils/imputeBOCF.R")
source("../../utils/imputeLOCF.R")
```

### Load data at 12-month

```{r}
# example: dat <- getData("../../../data/", 1)
# options: 1: "SummaryMonths1to4.csv"
#          2: "SummaryMonths1to12.csv"
#          3: "SummaryMonths5to12Weight.csv
#          4: "SummaryMonths5to12All.csv"
pathFile <- c("../../../data/")
numFile <- 2
dat <- getData(pathFile, numFile)
```

```{r}
names(dat)
```

### Data description

<!-- #region -->
The data variables used in our mediation analysis are described as follows:   
* *Treatment* is the treatment assignment variable, where it equals 0 for self-paced subjects and 1 for counselor-initiated subjects.  

* *WeightLoss12Month* is a log-ratio indicating the rate of weight change at four months with respect to the baseline weight. If the value is positive, it indicates a weight loss; if it is negative, it corresponds to gain, and 0 indicates no change between 12 months and baseline.

* *WeightLoss4Month* is a log-ratio indicating the rate of weight change at four months with respect to the baseline weight. If the value is positive, it indicates a weight loss; if it is negative, it corresponds to gain, and 0 indicates no change between 4 months and baseline. 

* The potential mediators that we plan to analyze are the *self-weighing* variable, the principal components derived from the electronically collected variables' PCA, and *WeightLoss4Month*. The logging variables quantify logging behavior in a number of days, and the caloric variables assess caloric intake and output. The logging and caloric variables are estimated over 12 months for the 12-month mediation analysis. The average caloric variables quantify the sum of the calories divided by the total number of logged days. For example, *AverageDailyBreakfastCalories* equals the sum of intake calories reported at breakfast divided by the total number of days when breakfast was logged.

   
| Logging            | Caloric                       |
|--------------------|-------------------------------|
| LoggedWeekendDays  | AverageDailyIntakeCalories    | 
| LoggedWeekDays     | AverageDailyExerciseCalories  |
| TotalLoggedDays    | AverageDailyBreakfastCalories |
| TotalFoodDays      | AverageDailyLunchCalories     |
| TotalExerciseDays  | AverageDailyDinnerCalories    |
| TotalBreakfastDays | AverageDailySnackCalories     |
| TotalLunchDays     |                               |
| TotalDinnerDays    |                               |
| TotalSnackDays     |                               |


<!-- #endregion -->

---
## Principal Component Analysis

Applying a PCA (principal component analysis) reduces the number of variables and adresses the multicollinearity of the electronically collected variables.



### Variables selection

```{r}
# Select the variable indexes to include in the Principal Component Analysis
myCols <-c("LoggedWeekDays", "LoggedWeekendDays", "TotalLoggedDays","TotalFoodDays",
            "TotalExerciseDays", 
            "TotalBreakfastDays", "TotalLunchDays", "TotalDinnerDays", "TotalSnackDays",
            "AverageDailyIntakeCalories", "AverageDailyExerciseCalories", 
            "AverageDailyBreakfastCalories",
            "AverageDailyLunchCalories", "AverageDailyDinnerCalories","AverageDailySnackCalories")
# Get the indexes of those variables
idxApp <- match(myCols, colnames(dat))
```

### Apply the PCA

```{r}
# Get the results of PCA
resPC <- getPCs(dat, idxApp, numPCs = 3) # max numPCs is 5
# Get the data frame including the first three principal components named app1, app2, app3.
dat <- resPC$data
```

```{r}
# Display the variables name
# print(names(dat))
# Select variables of interest
if (numFile == 1){
  myCols <-c("id","WeightLoss4Month", "Treatment", "Age", "NumWeigh", "App1", "App2", "App3")  
} else {
  myCols <-c("id","WeightLoss12Month", "WeightLoss4Month", "Treatment", "Age", "NumWeigh", "App1", "App2", "App3")      
}

myIdx <- match(myCols, colnames(dat))
dat <- dat[, myIdx]
print(names(dat))
```

### Display PCA results

```{r}
# resPC contains also the scree plot{scree.plot}, cumulative variance explained {CV.plot}, 
# the variables contribution of each component {contribCorr.plot()} and the quality of those 
# contributions {cos2Corr.plot()}.
options(repr.plot.width = 10, repr.plot.height = 5)
resPC$CV.plot
```

```{r}
options(repr.plot.width = 10, repr.plot.height = 7)
resPC$contribCorr.plot(1.5) # to get numbers use resPC$contribCorrNum.plot()
```

```{r}
options(repr.plot.width = 10, repr.plot.height = 7)
resPC$cos2Corr.plot(1.5) # to get numbers use resPC$cos2CorrNum.plot()
```

---
## Analysis


The first step of the mediation analysis is to establish if our candidates variables are mediators or not.    
**To analyze the mediation we observe the following flowchart:**

![alt text](images/Mediation_flowchart.svg "Mediation Analysis flowchart")

**We first estimated two linear regressions for both the mediator and the outcome. We used also a permutation test to assess any associations with respect to our samples.**


**Clean the data:** *impute with the last observation carried forward*

```{r}
dat <- imputeLOCF(dat, response = "both")
```

---
### Relationship association assessment

We will consider the following assumptions:

* App usage and Number of self-weighing are casually unrelated mediators.
* Observed confounder is : age. 


#### **Regression of weight loss on the treatment.**

```{r}
# Outcome model with confounders
out.fit <- lm(WeightLoss12Month ~ Treatment + Age, data = dat)
summary(out.fit)
```

#### **Permutation test: Weight loss ~ treatment assignment**

```{r}
options(repr.plot.width = 10, repr.plot.height = 5)
getPermut(dat$WeightLoss12Month, dat$Treatment, P = 10000, IVname = "Weight Loss")
```

***We confirmed that the treatment has a statistically significant predictive capability on weight loss outcome.***


#### Regression of App-use (*i.e.*, PC1 ) on the treatment.

```{r}
# Outcome model with confounders
out.fit <- lm(App1 ~ Treatment + Age, data = dat)
summary(out.fit)
```

#### **Permutation test: App-use ~ treatment assignment**

```{r}
getPermut(dat$App1, dat$Treatment, P = 10000, IVname = "App-use")
```

#### **Regression of App-calories (*i.e.*, PC2 ) on the treatment.**

```{r}
# Outcome model with confounders
out.fit <- lm(App2 ~ Treatment + Age, data = dat)
summary(out.fit)
```

#### **Permutation test: App-calories ~ treatment assignment**

```{r}
getPermut(dat$App2, dat$Treatment, P = 10000, IVname = "App-calories")
```

#### Regression of App-exercise (*i.e.*, PC3 ) on the treatment.

```{r}
# Outcome model with confounders
out.fit <- lm(App3 ~ Treatment + Age, data = dat)
summary(out.fit)
```

#### **Permutation test: App-exercise ~ treatment assignment**

```{r}
getPermut(dat$App3, dat$Treatment, P = 10000, IVname = "App-exercise")
```

#### **Regression of Self-weight on the treatment.**

```{r}
# Outcome model with confounders
out.fit <- lm(NumWeigh ~ Treatment + Age, data = dat)
summary(out.fit)
```

#### **Permutation test: Self-weight ~ treatment assignment**

```{r}
getPermut(dat$NumWeigh, dat$Treatment, P = 10000, IVname = "Self-weighing")
```

#### **Regression of 4-month weight loss on the treatment.**

```{r}
# Outcome model with confounders
out.fit <- lm(WeightLoss4Month ~ Treatment + Age, data = dat)
summary(out.fit)
```

#### **Permutation test: 4-month weight loss ~ treatment assignment**

```{r}
getPermut(dat$WeightLoss4Month, dat$Treatment, P = 10000, IVname = "4-month weight loss")
```

***We confirmed that the treatment has a statistically significant predictive capability on all potential mediators except for the app-calories and the app-exercise. We considered only app-use, self-weight and the 4-month weight loss variables for the mediation analysis for the 12 month-weight loss.***


---
## Mediation


**Our function `getMediationTable` to estimate the Average Causal Mediation Effect (ACME) and the Average Direct Effect (ADE) associated with each mediator is based on  the `mediation` library.**

```{r}
numFile <- 2
# rslts <- getMediationTable(dat, numFile, myRobustSE = TRUE, numSims = 10000, confLvl = 0.90) 
rslts <- getMediationTable(dat, numFile, myBoot = TRUE, numSims = 10000, confLvl = 0.90)
```

### **Mediation Results** 


```{r}
rslts$med.tbl %>% round(5)%>%
  kable("html") %>% 
  column_spec(c(1), bold = T, color = "black") %>%
  column_spec(c(2, 5, 8), bold = T, color = "steelblue") %>%
  column_spec(c(3, 4, 6, 7, 9, 10), bold = T, color = "sienna") %>%
  column_spec(c(11), bold = T, color = "black")%>%
  as.character() %>%
  display_html()
```

---
## Sensitivity Analysis

```{r}
# Sensitivity for the 4-month weight loss
sens4Month.out <- medsens(rslts$med.rslts$med4Month.out, rho.by = 0.01, effect.type = "indirect", sims = 10000)
```

```{r}
# Sensitivity for the app-use
sensApp.out <- medsens(rslts$med.rslts$medApp.out, rho.by = 0.01, effect.type = "indirect", sims = 10000)
# Sensitivity for the self-weight
sensNumW.out <- medsens(rslts$med.rslts$medNumW.out, rho.by = 0.01, effect.type = "indirect", sims = 10000)
```

### 4-month weight loss


#### Results

```{r}
summary(sens4Month.out)
```

#### Sensitivity with respect to error correlation

```{r}
options(repr.plot.width = 10, repr.plot.height = 7)
plot(sens4Month.out,  ylim = c(-0.18, 0.18), main  =  expression(paste("App-use ACME(", rho, ")")))
```

The above figure shows the estimated true values of ACMEs as a function of the sensitivity parameter $\rho$. The parameter $\rho$
expresses the correlation between the error terms in the mediator and the outcome models. The solid thick line and its shaded area represent the ACME's point estimates and their 90% confidence intervals, respectively.    
According to this analysis, the ACME is estimated to be negative when  $\rho$ is larger than 0.56, and the ACME is statistically indistinguishable from zero at the 90% level when the parameter $\rho$ is larger than 0.48 and less than 0.62.


#### Sensitivity with respect to proportion of variance explained

```{r}
plot(sens4Month.out, sens.par = "R2", r.type = "total", sign.prod = "positive",
     xlim = c(-0.0, 0.8), ylim = c(-0.0, 0.65))
```

In the sensitivity analyses plot above, the ACME estimates are plotted against $\tilde{R}^2_Y \tilde{R}^2_M$, the proportions of the total variance in the outcome and mediator variables, respectively, that would be explained by a hypothetical unobserved pretreatment confounder.
The bold line represents the various product combinations in terms of the coefficient of determination where the ACME would become 0. In this case, the estimated ACME will be 0 if the product of $\tilde{R}^2_M$ and $\tilde{R}^2_Y$ is 0.1621. For example, this implies that an unobserved pretreatment confounder would explain 20% of the variation in the participants’ app-use and 81.05% of the variation in their weight loss.


### App-use


#### Results

```{r}
summary(sensApp.out)
```

#### Sensitivity with respect to error correlation

```{r}
options(repr.plot.width = 10, repr.plot.height = 7)
plot(sensApp.out,  ylim = c(-0.18, 0.18), main  =  expression(paste("App-use ACME(", rho, ")")))
```

The above figure shows the estimated true values of ACMEs as a function of the sensitivity parameter $\rho$. The parameter $\rho$
expresses the correlation between the error terms in the mediator and the outcome models. The solid thick line and its shaded area represent the ACME's point estimates and their 90% confidence intervals, respectively.    
According to this analysis, the ACME is estimated to be negative when  $\rho$ is larger than 0.25, and the ACME is statistically indistinguishable from zero at the 90% level when the parameter $\rho$ is larger than 0.14 and less than 0.35.


#### Sensitivity with respect to proportion of variance explained

```{r}
plot(sensApp.out, sens.par = "R2", r.type = "total", sign.prod = "positive",
     xlim = c(-0.0, 0.6), ylim = c(-0.0, 0.85))
```

In the sensitivity analyses plot above, the ACME estimates are plotted against $\tilde{R}^2_Y \tilde{R}^2_M$, the proportions of the total variance in the outcome and mediator variables, respectively, that would be explained by a hypothetical unobserved pretreatment confounder.
The bold line represents the various product combinations in terms of the coefficient of determination where the ACME would become 0. In this case, the estimated ACME will be 0 if the product of $\tilde{R}^2_M$ and $\tilde{R}^2_Y$ is 0.0326. For example, this implies that an unobserved pretreatment confounder would explain 20% of the variation in the participants’ app-use and 16.3 of the variation in their weight loss.



### Self-weight


#### Results

```{r}
summary(sensNumW.out)
```

#### Sensitivity with respect to error correlation

```{r}
plot(sensNumW.out,  ylim = c(-0.18, 0.18), main  =  expression(paste("Self-weight ACME(", rho, ")")))
```

The above figure shows the estimated true values of ACMEs as a function of the sensitivity parameter $\rho$. The parameter $\rho$
expresses the correlation between the error terms in the mediator and the outcome models. The solid thick line and its shaded area represent the ACME's point estimates and their 90% confidence intervals, respectively.    
According to this analysis, the ACME is estimated to be negative when  $\rho$ is larger than 0.34, and the ACME is statistically indistinguishable from zero at the 90% level when the parameter $\rho$ is larger than 0.24 and less than 0.44.


### Sensitivity with respect to proportion of variance explained

```{r}
plot(sensNumW.out, sens.par = "R2", r.type = "total", sign.prod = "positive",
     xlim = c(-0.0, 0.7), ylim = c(-0.0, 0.7))
```

In the sensitivity analyses plot above, the ACME estimates are plotted against $\tilde{R}^2_Y \tilde{R}^2_M$, the proportions of the total variance in the outcome and mediator variables, respectively, that would be explained by a hypothetical unobserved pretreatment confounder.
The bold line represents the various product combinations in terms of the coefficient of determination where the ACME would become 0. In this case, the estimated ACME will be 0 if the product of $\tilde{R}^2_M$ and $\tilde{R}^2_Y$ is 0.0779. For example, this implies that an unobserved pretreatment confounder would explain 20% of the variation in the participants’ app-use and 38.95% of the variation in their weight loss.


## Conclusion


Overall, the results suggest that under the assumption of independence between mechanisms, the causal mediation effects are positive and not significant for the app-use and it is significant for the self weight and for the 4-month weight loss mediator, and the estimates for the 4-month weight loss showed robustness to the
possible unobserved pretreatment mediator–outcome confounding to varying degrees. However, the estimates for app-use are not robust to the
possible unobserved pretreatment mediator–outcome confounding to varying degrees. The results of the mediation analysis confirmed that there are no sufficient evidence to detect a mediation effect from the app-calories variable.   

```{r}
sessionInfo()
```
