---
title: "Fit Blue Data Description"
output:   
  html_document:
    fig_width: 10
    fig_height: 8
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    smooth_scroll: true
    theme: cerulean
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

#Read libraries

library(knitr)
library(tidyverse)
library(dplyr)
library(reshape2)
library(lubridate)
library(kableExtra)
library(gridExtra)
library(ggrepel)
library(DiagrammeR)
library(caret)
library(FactoMineR)
library(factoextra)
library(fastDummies)
library(corrplot)
library(pROC)
library(ggthemes)

#Read external functions
source("../utils/demoFunc.R")
source("../utils/analysisFunc.R")

#Read File

# Get input folder and data file
dirInput <- "../../data/"
inputFile <- paste(dirInput, "ProcessedData_2019_04_03.txt", sep = "")
# Read input file 
processed.data <- read.table(file = inputFile, header = TRUE, quote = "\"", sep = "\t", fill = TRUE)



#Take out first column with just row number
processed.data <- processed.data %>%
  select(-X)

#Put into r date form
processed.data$VST_DATE_Base <- mdy(processed.data$VST_DATE_Base)
processed.data$loseit_date <- mdy(processed.data$loseit_date)

#Make Success a factor
processed.data$Success <- factor(processed.data$Success)

#Make Treatment a factor and change Treatment names
processed.data$tx <- factor(processed.data$tx)
levels(processed.data$tx) <- c("Counselor-Initiated", "Self-Paced")

# Get input files for Summary Weeks
# Get input folder and data file
inputFile <- c("../../data/SummaryWeeks1to4.csv",
               "../../data/SummaryWeeks1to8.csv")

# Read input file 
SummaryWeeks1to4.data <- read.csv(file = inputFile[1], header = TRUE)
SummaryWeeks1to8.data <- read.csv(file = inputFile[2], header = TRUE)

 
###################################################################################
# temp: change  weightloss such positive is weight loss and negative is weight gain
# Assign 0 for controlled and 1 for treated.
SummaryWeeks1to8.data$Treatment[SummaryWeeks1to8.data$Treatment == 2] <- 0
SummaryWeeks1to4.data$Treatment[SummaryWeeks1to4.data$Treatment == 2] <- 0
# # Change sign for weight loss so that: positive is a loss and negative is a gain.
# SummaryWeeks1to8.data$WeightLoss4Month <- -SummaryWeeks1to8.data$WeightLoss4Month
# SummaryWeeks1to8.data$WeightLoss12Month <- -SummaryWeeks1to8.data$WeightLoss12Month

# Change sign for intake calorie so that: intake calories are negative and burnt calories are positive
SummaryWeeks1to8.data$AvgTotalCal <- -SummaryWeeks1to8.data$AvgTotalCal
SummaryWeeks1to8.data$AvgBreakfastCal <- -SummaryWeeks1to8.data$AvgBreakfastCal
SummaryWeeks1to8.data$AvgLunchCal <- -SummaryWeeks1to8.data$AvgLunchCal
SummaryWeeks1to8.data$AvgDinnerCal <- -SummaryWeeks1to8.data$AvgDinnerCal
SummaryWeeks1to8.data$AvgSnackCal <- -SummaryWeeks1to8.data$AvgSnackCal

# Change sign for weight loss so that: positive is a loss and negative is a gain.
SummaryWeeks1to4.data$WeightLoss4Month <- SummaryWeeks1to4.data$WeightLoss4Month
SummaryWeeks1to4.data$WeightLoss12Month <- SummaryWeeks1to4.data$WeightLoss12Month

# Change sign for intake calorie so that: intake calories are negative and burnt calories are positive
SummaryWeeks1to4.data$AvgTotalCal <- -SummaryWeeks1to4.data$AvgTotalCal
SummaryWeeks1to4.data$AvgBreakfastCal <- -SummaryWeeks1to4.data$AvgBreakfastCal
SummaryWeeks1to4.data$AvgLunchCal <- -SummaryWeeks1to4.data$AvgLunchCal
SummaryWeeks1to4.data$AvgDinnerCal <- -SummaryWeeks1to4.data$AvgDinnerCal
SummaryWeeks1to4.data$AvgSnackCal <- -SummaryWeeks1to4.data$AvgSnackCal

#Take out first column with X
SummaryWeeks1to4.data$X <- NULL
SummaryWeeks1to8.data$X <- NULL
```

# Abstract

Participants of the Fit Blue study, a weight loss study implemented at the Lackland Air Force Base, were asked to record daily diet and activity on the Lose It! application. Our objective was to investigate the effects of electronic self-monitoring behavior (using the commercial Lose It!TM application) and weight-loss interventions (with differing amounts of counselor feedback and support) on 4 and 12-month weight loss. We compared the results from two interventions of a randomized controlled trial. Counselor-initiated (CI) treatment participants received early, consistent support from interventionists, and self-paced (SP) treatment participants received assistance on request. Participants (n= 191) who were active duty military personnel were encouraged to self-monitor diet and exercise with the Lose It!TM application or website. We examined logging trends throughout the study for associations between intervention assignment and app usage. We conducted a mediation analysis of the intervention assignment on weight loss through multiple mediators: app usage (calculated from the first principal component of electronically collected variables), number of weigh-ins and 4-month weight change. We used linear regression to predict weight loss at 4 and 12 months; accuracy was measured using cross-validation. Since the app provides many variables, We performed principal components analysis to reduce data dimension and address collinearity.

# Descriptive Analysis

## Demographics

All particpants in the study were on active duty in the military and had to have at least one year left at the Air Force base so that they would likely be available in person for the 4 and 12 month follow-ups. The requirements for participation were as follows: 18 years or older, BMI greater or equal to 25.0, access to computer and email, and cleared to participate by a healthcare provider.
      

There were 212 total participants in this study. Only 191 records were successfully pulled from the Lose It! database.

```{r, echo = FALSE}
# Create data frame for demographics
inputFile <- paste("../../data/ScreeningBMI.csv", sep = "")
# Read input file ignoring all single quotes
screening.data <- read.csv(file = inputFile, header = TRUE)

new.data <- merge(processed.data, screening.data, by = "id")

demographics <- new.data %>%
  group_by(id) %>%
  summarise(Age = AGE_yrs[1], Gender = GENDER[1], Race = race3[1], 
            Education = educcat[1], InitialBMI = bmi_sv[1], BMI_Group = getBMIGroup(bmi_sv[1]),
            tx = tx[1],.groups = 'drop')

# Create data frame for demographics for Counselor-Initiated
demographicsCI <- demographics %>%
  filter(tx == "Counselor-Initiated")

# Create data frame for demographics for Self-Paced
demographicsSI <- demographics %>%
  filter(tx == "Self-Paced") 


#Initialize demographic summary table
demo.summary1 <- data.frame(matrix(ncol = 3, nrow = 10))
rownames(demo.summary1) <- c("Female", "Male", "African American", "Caucasian American", "Other", "Less than College",
                            "College or Above", "Overweight (BMI: 25-29.9)", 
                            "Obese (BMI: 30+)", "Age (mean, sd)")
colnames(demo.summary1) <- c("Number of Participants CI", "Number of Participants SP", "Total(%)")

#Percentage of characteristic in study total for Counselor-Initiated
demo.summary1[1,1] <- demo(demographicsCI$Gender, "Female") 
demo.summary1[2,1] <- demo(demographicsCI$Gender, "Male") 
demo.summary1[3,1] <- demo(demographicsCI$Race, "African American") 
demo.summary1[4,1] <- demo(demographicsCI$Race, "Caucasian American") 
demo.summary1[5,1] <- demo(demographicsCI$Race, "Others") 
demo.summary1[6,1] <- demo(demographicsCI$Education, "Less than College") 
demo.summary1[7,1] <- demo(demographicsCI$Education, "College or Above") 
demo.summary1[8,1] <- demo(demographicsCI$BMI_Group, "Overweight") 
demo.summary1[9,1] <- demo(demographicsCI$BMI_Group, "Obese")
demo.summary1[10,1] <- paste(round(mean(demographicsCI$Age),1),"(",
                             round(sd(demographicsCI$Age),1), ")", sep= "")


#Percentage of characteristic in study total for Self-Initiated
demo.summary1[1,2] <- demo(demographicsSI$Gender, "Female") 
demo.summary1[2,2] <- demo(demographicsSI$Gender, "Male") 
demo.summary1[3,2] <- demo(demographicsSI$Race, "African American") 
demo.summary1[4,2] <- demo(demographicsSI$Race, "Caucasian American") 
demo.summary1[5,2] <- demo(demographicsSI$Race, "Others") 
demo.summary1[6,2] <- demo(demographicsSI$Education, "Less than College") 
demo.summary1[7,2] <- demo(demographicsSI$Education, "College or Above") 
demo.summary1[8,2] <- demo(demographicsSI$BMI_Group, "Overweight") 
demo.summary1[9,2] <- demo(demographicsSI$BMI_Group, "Obese") 
demo.summary1[10,2] <- paste(round(mean(demographicsSI$Age),1),"(",
                             round(sd(demographicsSI$Age),1), ")", sep= "")

#Calculate totals
demo.summary1[1,3] <- percentdemo(demographics$Gender, "Female") 
demo.summary1[2,3] <- percentdemo(demographics$Gender, "Male") 
demo.summary1[3,3] <- percentdemo(demographics$Race, "African American") 
demo.summary1[4,3] <- percentdemo(demographics$Race, "Caucasian American") 
demo.summary1[5,3] <- percentdemo(demographics$Race, "Others") 
demo.summary1[6,3] <- percentdemo(demographics$Education, "Less than College") 
demo.summary1[7,3] <- percentdemo(demographics$Education, "College or Above") 
demo.summary1[8,3] <- percentdemo(demographics$BMI_Group, "Overweight") 
demo.summary1[9,3] <- percentdemo(demographics$BMI_Group, "Obese") 
demo.summary1[10,3] <- paste(round(mean(demographics$Age),1),"(",
                             round(sd(demographics$Age),1), ")", sep= "")

# calculate percentage for demographics
df <- demo.summary1[-10,-3]
df[,1] <- as.numeric(df[,1])
df[,2] <- as.numeric(df[,2])
df$Total <- df[,1] + df[,2]
df$CIperc<- round(100*(df[,1]/df$Total),1)
df$SIperc<- round(100*(df[,2]/df$Total),1)

# calculate P-value
chsqtestGender <- chisq.test(x = df[1:2,c(1,2)])
chsqtestRace <- chisq.test(x = df[3:5,c(1,2)])
chsqtestEducation <- chisq.test(x = df[6:7,c(1,2)])
chsqtestBMI <- chisq.test(x = df[8:9,c(1,2)])
ttestAge <- t.test(Age~Treatment, data = SummaryWeeks1to4.data)
pval <- c(chsqtestGender$p.value, chsqtestRace$p.value, chsqtestEducation$p.value, 
          chsqtestBMI$p.value, ttestAge$p.value) %>% round(3) %>% as.character()


# add percentage to the table  
vecTmp <- str_c("(", df[,4], "%)")
demo.summary1[-10,1] <- str_c(demo.summary1[-10,1],vecTmp)

vecTmp <- str_c("(", df[,5], "%)")
demo.summary1[-10,2] <- str_c(demo.summary1[-10,2],vecTmp)

# add P-values
demo.summary1$`P-value` <- c(pval[1], "", "", pval[2],"", pval[3], "", pval[4], "", pval[5])


kable(demo.summary1) %>%
  kable_styling(c("striped", "bordered")) %>%
  group_rows("Gender", 1, 2) %>%
  group_rows("Race", 3, 5) %>%
  group_rows("Education", 6,7) %>%
  group_rows("BMI Category", 8,9) %>%
  group_rows("Age", 10, 10)
```





## Average % Days Logged-in

Login Days decreased steadily over the course of the study. Weekend and Weekday logins tended to follow the same trends with peaks and falls in the average logins. The increase at week 46 aligns with the participants' receivals of reminder cards about the upcoming 12 month weigh-in

```{r}
#Average % Days logged-in per week over all 52 weeks of the study broken down by weekend/weekday

logins.weekly <- processed.data %>%
  mutate(Weekday = weekdays(loseit_date, abbreviate = TRUE)) %>%
  filter(Week != 53) %>%
  group_by(id, Week, tx, GENDER) %>% 
  summarize(PerWeekdays = sum(Weekday %in% c("Mon", "Tue", "Wed", "Thu", "Fri"))/5,
            PerWeekendDays = sum(Weekday %in% c("Sat", "Sun"))/2, .groups = 'drop')

logins.weekly.all <- logins.weekly %>%
  group_by(Week) %>%
  summarize(week = mean(PerWeekdays), weekend = mean(PerWeekendDays), .groups = 'drop')

logins.weekly.all <- melt(logins.weekly.all, id.vars = c("Week"), variable.name = "Day")

ggplot(logins.weekly.all, aes(x = Week)) +
  geom_line(aes(y = value, col = Day), size = 1) +
  labs(x = "Week", y = "Average % Days Logged-in") +
  scale_x_continuous(breaks = seq(0,52,2)) +
  scale_y_continuous(breaks = seq(.5,1,.05)) +
  theme(legend.position = c(.1,.1), legend.background = element_rect(color = "black", linetype = "solid"))
```


## Number of Login Days in the First 4 Weeks

Trends within treatment groups are apparent. Counselor-Initiated groups have similar login habits in the first 4 weeks while Self-Paced groups have more similar login habits regardless of Success. Logins are just starting to decrease in the unsuccessful Counselor-Initiated group in the fourth week which led us to believe there may be a reason to explore the first 8 weeks of the study. Another point to note is that consistent logins in the first 4 weeks didn't seem to be a predictor of successful weight loss for Self-Paced people.

```{r, echo = FALSE}
#Create a data frame with average calories and number of logins per week for each ID number

weeks4 <- processed.data %>%
  filter(Week %in% 1:4) %>%
  group_by(id, loseit_date) %>%
  select(id, loseit_date, tx, Success, Week) %>%
  group_by(id, Week) %>%
  summarize(tx = tx[1], Success= Success[1], NumOfEntries = n(), .groups = 'drop')

#Create dataframe with row for each id's weeks 1-4
allID <- unique(weeks4$id)
allID4Week <- as.data.frame(rep(allID, 4))
colnames(allID4Week) <- "id"
allID4Week <- arrange(allID4Week, id)
allID4Week$Week <- rep(seq(1:4))

#merge dataframes
allWeeks <- merge(allID4Week, weeks4, all.x = TRUE)
allWeeks$NumOfEntries <- as.numeric(allWeeks$NumOfEntries)

#Fill treatment and Success data down
allWeeks <- allWeeks %>%
  group_by(id) %>%
  fill(tx, .direction = "down") %>%
  fill(Success, .direction = "down") 
  
allWeeks[is.na(allWeeks)] <- 0

levels(allWeeks$Success) <- c("Not Successful", "Successful")

#Stacked bar graph of the number of Logins in the first 8 weeks

allWeeks$NumOfEntries <- factor(allWeeks$NumOfEntries)

cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

ggplot(allWeeks, aes(x = factor(Week), fill = NumOfEntries)) +
  geom_bar(position = "fill") +
  facet_grid(.~ Success + tx) +
  #scale_fill_manual(values = cbbPalette, name = "Number of days\nLogged per Week") +
  #scale_color_colorblind(name = "Number of days\nLogged per Week")+
  labs(y = "Frequency", x = "Week") +
  scale_fill_brewer(palette = "Spectral",name = "Number of days\nLogged per Week")


```

## Number of Login Days in the first 8 weeks

* When looking at the first 8 weeks, it is easier to see the trends among groups. The Counselor-initiated Success group still looks to be the most stable in terms of days logged in

* The Self-Paced Success group began to level out after the 4th week whereas the unsuccessful Self-Paced group's logins continued to decrease

* The Counselor-Initiated unsuccessful group remained high in logins but there was more variability in the number of logins between weeks.

```{r, echo = FALSE}
#Create a data frame with average calories and number of logins per week for each ID number

weeks8 <- processed.data %>%
  filter(Week %in% 1:8) %>%
  group_by(id, loseit_date) %>%
  select(id, loseit_date, tx, Success, Week) %>%
  group_by(id, Week) %>%
  summarize(tx = tx[1], Success= Success[1], NumOfEntries = n(), .groups = 'drop')

#Create dataframe with row for each id's weeks 1-4
allID <- unique(weeks8$id)
allID8Week <- as.data.frame(rep(allID, 8))
colnames(allID8Week) <- "id"
allID8Week <- arrange(allID8Week, id)
allID8Week$Week <- rep(seq(1:8))

#merge dataframes
allWeeks <- merge(allID8Week, weeks8, all.x = TRUE)
allWeeks$NumOfEntries <- as.numeric(allWeeks$NumOfEntries)

#Fill treatment and Success data down
allWeeks <- allWeeks %>%
  group_by(id) %>%
  fill(tx, .direction = "down") %>%
  fill(Success, .direction = "down") 
  
allWeeks[is.na(allWeeks)] <- 0

levels(allWeeks$Success) <- c("Not Successful", "Successful")

#Stacked bar graph of the number of Logins in the first 8 weeks

allWeeks$NumOfEntries <- factor(allWeeks$NumOfEntries)

cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

ggplot(allWeeks, aes(x = factor(Week), fill = NumOfEntries)) +
  geom_bar(position = "fill") +
  facet_grid(.~ Success + tx) +
  #scale_fill_manual(values = cbbPalette, name = "Number of days\nLogged per Week") +
  #scale_color_colorblind(name = "Number of days\nLogged per Week")+
  labs(y = "Frequency", x = "Week") +
  scale_fill_brewer(palette = "Spectral",name = "Number of days\nLogged per Week")

#theme(legend.position = "top", legend.direction = "horizontal") + guides(fill = guide_legend(nrow = 1), override.aes = list(size = 4))



#Save as .tiff
#ggsave("NumOfEntries.tiff", units="in", dpi=600, compression = 'lzw')

#Save as .png
# ggsave("NumOfEntries.png", width = 6, height = 3.5,units="in", dpi=600)
```
   
   
## Number of Login Days in 4 months
   
```{r, echo = FALSE}
#Create a data frame with average calories and number of logins per week for each ID number

weeks16 <- processed.data %>%
  filter(Week %in% 1:16) %>%
  group_by(id, loseit_date) %>%
  select(id, loseit_date, tx, Success, Week) %>%
  group_by(id, Week) %>%
  summarize(tx = tx[1], Success= Success[1], NumOfEntries = n(), .groups = 'drop')

#Create dataframe with row for each id's weeks 1-16
allID <- unique(weeks16$id)
allID16Week <- as.data.frame(rep(allID, 16))
colnames(allID16Week) <- "id"
allID16Week <- arrange(allID16Week, id)
allID16Week$Week <- rep(seq(1:16))

#merge dataframes
allWeeks <- merge(allID16Week, weeks16, all.x = TRUE)
allWeeks$NumOfEntries <- as.numeric(allWeeks$NumOfEntries)

#Fill treatment and Success data down
allWeeks <- allWeeks %>%
  group_by(id) %>%
  fill(tx, .direction = "down") %>%
  fill(Success, .direction = "down") 
  
allWeeks[is.na(allWeeks)] <- 0

levels(allWeeks$Success) <- c("Not Successful", "Successful")

#Stacked bar graph of the number of Logins in the first 8 weeks

allWeeks$NumOfEntries <- factor(allWeeks$NumOfEntries)

cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

ggplot(allWeeks, aes(x = factor(Week), fill = NumOfEntries)) +
  geom_bar(position = "fill") +
  facet_grid(.~ tx) +
  #scale_fill_manual(values = cbbPalette, name = "Number of days\nLogged per Week") +
  #scale_color_colorblind(name = "Number of\nEntries")+
  labs(y = "Frequency", x = "Week") +
  scale_fill_brewer(palette = "Spectral",name = "Number of days\nLogged per Week")

#theme(legend.position = "top", legend.direction = "horizontal") + guides(fill = guide_legend(nrow = 1), override.aes = list(size = 4))



#Save as .tiff
#ggsave("NumOfEntries.tiff", units="in", dpi=600, compression = 'lzw')

#Save as .png
# ggsave("NumOfEntries.png", width = 6, height = 3.5,units="in", dpi=600)
```

